---
title: "Standard report for the 16S amplicon sequencing"
author: "Diener Lab 16S pipeline"
date: now
format:
    html:
        embed-resources: true
        html-math-method: katex
        code-fold: true
        code-summary: "Show the code"
        code-line-numbers: true
        df-print: paged
---

```{r}
library(miso)
library(tidyverse)
rdata = "r_data"

AMPLEN = 806 - 515

theme_minimal() |> theme_set()
```

# General statistics

Let's read in the quality control results and check some basics.

```{r}
qc <- readRDS(file.path(rdata, "qc.rds"))
```

## Error plots

We start by looking at the qualities. We cut at 290 bp as this is the amplicon size.

```{r}
qc$quality_plot + xlim(0, AMPLEN)
```

## Entropy

This is to checkl that there are no adapters there anymore. This should go up across the reads.

```{r}
qc$entropy_plot + xlim(0, AMPLEN)
```

## Length checks

Finally we check for the lengths of the amplicons. Shorter or longer amplicons are a sign of unspecific amplification
or contamination with mitochondria or chloroplasts.

```{r}
qc$length_plot +
  scale_fill_continuous(palette=viridisLite::magma(100)) +
  geom_vline(xintercept=AMPLEN, linetype="dashed", color="yellow", linewidth=1)
```

# Basic analysis

```{r}
ps <- readRDS(file.path(rdata, "phyloseq_with_tree.rds"))
man <- readxl::read_excel(Sys.glob("raw/*.xlsx"), skip=9) |>
  mutate(Barcode = str_split_i(Barcode, " ", 2), sample_id = `Externe ID`) |>
  drop_na(sample_id) |>
  mutate(type = c("sample", "control")[str_detect(sample_id, "pos|neg") + 1])
sdata <- sample_data(ps) |> as("data.frame")
sdata <- man |> inner_join(sdata, by=join_by(Barcode==id)) |> as.data.frame()
rownames(sdata) <- sdata[["Barcode"]]
sample_data(ps) <- sdata
sample_names(ps) <- sample_data(ps)[["sample_id"]]
```

```{r}
sample_sums(ps) |> enframe("sample", "library_size")
```

Convert abundances to long tables for ASVs and Genera.

```{r}
for (ra in rank_names(ps)) {
  plot_taxa(ps, level=ra)
  ggsave(paste0("figures/", tolower(ra), "_top12.png"), dpi=300, width=8, height=6)
}

```

```{r}
asvs <- taxa_count(ps, lev=NA)[reads > 0][order(sample, -reads, taxa)]

g <- taxa_count(ps, lev="genus", zeros=T)[order(sample, -reads, taxa)]
mat <- as.matrix(g)

write_csv(asvs, "tables/asv_abundances.csv")
write_csv(g, "tables/genus_abundances.csv")
asvs
```

```{r, fig.width=8, fig.height=24}
library(ComplexHeatmap)

hm <- Heatmap(
  log10(mat+0.5)[sort(rownames(mat)),] |> t(),
  col=viridisLite::magma(256),
  row_names_max_width = unit(12, "cm"),
  name="log10(reads)",
  column_dend_side="bottom",
  column_names_side="top",
  column_names_rot=45
)

png("figures/genus_heatmap.png", width=8, height=24, unit="in", res=300)
print(hm)
dev.off()

hm
```

# Ordination

Quick check how samples differ.

```{r}
ord <- ordinate(ps, method="PCoA", distance="bray")
```

```{r, fig.width=6, fig.height=4}
plot_ordination(ps, ord, label="sample_id", color="type")
ggsave("figures/pcoa.png", dpi=300, width=6, height=4)
```
