---
title: "Standard report for the 16S amplicon sequencing"
author: "Diener Lab 16S pipeline"
date: now
format:
    html:
        embed-resources: true
        html-math-method: katex
        code-fold: true
        code-tools: true
        code-line-numbers: true
        df-print: paged
---

```{r}
library(miso)
library(tidyverse)
rdata = "r_data"

AMPLEN = 806 - 515

theme_minimal() |> theme_set()
```

# General statistics

Let's read in the quality control results and check some basics.

```{r}
qc <- readRDS(file.path(rdata, "qc.rds"))
```

## Error plots

We start by looking at the qualities. We cut at 290 bp as this is the amplicon size.

```{r}
qc$quality_plot + xlim(0, AMPLEN)
```

## Entropy

This is to checkl that there are no adapters there anymore. This should go up across the reads.

```{r}
qc$entropy_plot + xlim(0, AMPLEN)
```

## Length checks

Finally we check for the lengths of the amplicons. Shorter or longer amplicons are a sign of unspecific amplification
or contamination with mitochondria or chloroplasts.

```{r}
qc$length_plot +
  scale_fill_continuous(palette=viridisLite::magma(100)) +
  geom_vline(xintercept=AMPLEN, linetype="dashed", color="yellow", linewidth=1)
```

We get a more detailed overview of amplicon length variations in the amplicon types. We will visualize those for each sample along with the library size.

```{r}
amplis <- read_csv("amplicon_types.csv") |>
  pivot_longer(
    cols = c(target, mitochondrial, chloroplast, short, long),
    names_to="type",
    values_to="reads"
  )

ggplot(amplis) +
  aes(x=id, y=reads, fill=type) +
  geom_col(position="dodge2") +
    geom_hline(yintercept=c(1e3, 10e3, 100e3), linetype="dashed") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  scale_y_sqrt(breaks=c(0, 10^(2:7))) +
  labs(x="", y="#reads")

ggsave("figures/amplicon_types.png", width=8, height=6, dpi=300)
```

# Basic analysis

```{r}
ps <- readRDS(file.path(rdata, "phyloseq_with_tree.rds"))
```

```{r}
sample_sums(ps) |> enframe("sample_id", "library_size")
```

Convert abundances to long tables for ASVs and Genera.

```{r}
for (ra in rank_names(ps)) {
  plot_taxa(ps, level=ra)
  ggsave(paste0("figures/", tolower(ra), "_top12.png"), dpi=300, width=8, height=6)
}

```

```{r}
asvs <- taxa_count(ps, lev=NA)[reads > 0][order(sample, -reads, taxa)]

g <- taxa_count(ps, lev="genus", zeros=T)[order(sample, -reads, taxa)]
mat <- as.matrix(g)

write_csv(asvs, "tables/asv_abundances.csv")
write_csv(g[reads > 0], "tables/genus_abundances.csv")
asvs
```

```{r, fig.width=24, fig.height=8}
library(ComplexHeatmap)

hm <- Heatmap(
  log10(mat+0.5)[sort(rownames(mat)),],
  col=viridisLite::magma(256),
  column_names_max_height = unit(12, "cm"),
  name="log10(reads)",
  row_dend_side="right",
  row_names_side="left",
  column_names_rot=45
)

png("figures/genus_heatmap.png", width=24, height=8, unit="in", res=300)
print(hm)
dev.off()

hm
```

# Ordination

Quick check how samples differ.

```{r}
ord <- ordinate(ps, method="PCoA", distance="bray")
```

```{r, fig.width=6, fig.height=4}
plot_ordination(ps, ord, label="id", color="type")
ggsave("figures/pcoa.png", dpi=300, width=6, height=4)
```
